Improving Haskell Theory Exploration

### 2015-11-23 TransferReport/background.tex
More TODOs

\iffalse TODO: Not immediately clear how the problem of reducing the set of expressions relates to premise selection \fi
\iffalse TODO: This detailed comparison with premise selection is opaque without definiing/explaining what premise selection is \fi

@@

Both of these approaches, premise selection and clustering, use \emph{machine learning} (ML) algorithms to analyse (logical or software) expressions, and hence rely on \emph{feature extraction} to transform the data into a suitable representation and remove irrelevant details. This has two benefits:

\begin{itemize}
  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}. This reduces the size of the machine learning problem, improving efficiency (e.g. running time).
  \item We avoid learning irrelevant details, such as the text encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
\end{itemize}

@@

However, this is too optimistic. Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits. Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.

@@

There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor. Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning! \footnote{Consider a classification problem, to assign a label $l \in L$ to each input. If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}

@@

>>>>

\iffalse TODO: Not immediately clear how the problem of reducing the set of expressions relates to premise selection \fi
\iffalse TODO: This detailed comparison with premise selection is opaque without defining/explaining what premise selection is \fi

@@

Both of these approaches, premise selection and clustering, use \emph{machine learning} (ML) algorithms to analyse (logical or software) \iffalse TODO: software vs logical expr? \fi expressions, and hence rely on \emph{feature extraction} to transform the data into a suitable representation and remove irrelevant details. This has two benefits:

\begin{itemize}
  \item \emph{Feature vectors} (ordered lists of features) are chosen to be more compact \iffalse TODO: unclear \fi than the data they're extracted from: feature extraction is \emph{lossy compression}. This reduces the size of the machine learning problem, improving efficiency (e.g. running time).
  \item We avoid learning irrelevant details, such as the text encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
\end{itemize}

@@

However, this is too optimistic. Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits. \iffalse TODO: It depends on your feature extraction, and why bits? \fi Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
\iffalse TODO: I do not see this has to be this way\fi

@@

There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor. Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning! \iffalse TODO: not sure that's the right argument for a more-or-less common observation that ML & FE are coming together \fi \footnote{Consider a classification problem, to assign a label $l \in L$ to each input. If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}

@@

\iffalse TODO: I have had a look now at section 5. I do not think feature extraction and clustering is explained/defined there, either. Seeing your contribution is feature extraction algorithms, you need to define feature vectors and clusters *here* \fi

### 2015-11-23 TransferReport/contributions.tex
Added more feedback
More TODOs

\section{Contributions}
\label{contributions}

\subsection{Recurrent Clustering}
\label{recurrentclustering}

@@

Recurrent clustering tackles this problem in a straightforward way: we look up the expressions \emph{referenced by} each identifier, and use \emph{their} similarity in place of the identifiers'. This gives rise to a recursive process, where feature extraction is interleaved with multiple rounds of clustering. For this recursive process to be well-founded, we impose a topological ordering on expressions based on their dependencies (the expressions they reference). This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and several mutually-recursive expressions may appear at the same level.

\subsubsection{Algorithm}
\label{algorithm}

@@

    recurrentCluster :: [[(Global, Expr)]] -> [[Global]]
    recurrentCluster = go ([],[])
      where go (fs, db) []       = db
            go (fs, db) (es:ess) = let fs' = fs ++ map (extract db) es
                                       db' = kMeans fs'
                                    in go (fs', db') ess
            extract db (i, e) = (i, rt db e)
  \end{lstlisting}
  \caption{Feature extraction for Haskell Core. We send named expressions to \hs{recurrentCluster}, topologically sorted: if an element contains multiple \hs{(Global, Expr)} pairs, they are mutually-recursive. Syntax trees are first converted into \hs{RoseTree}s of \hs{Feature}s, these trees are converted to matrices, then flattened into feature vectors, then \hs{kMeans} clusters the feature vectors. The result is a list of clusters, naming their elements. More details are given in \S \ref{algorithm}.}
  \label{featureextractionalgorithm}
\end{figure}

Our feature extraction algorithm is given in figure \ref{featureextractionalgorithm}. Here we highlight some key features:

>>>>

\section{Contributions}
\label{sec:contributions}

\subsection{Recurrent Clustering}
\label{sec:recurrentclustering}

@@

Recurrent clustering tackles this problem in a straightforward way: \iffalse TODO: What would be a non-straightforward way? \fi we look up the expressions \emph{referenced by} each identifier, and use \emph{their} similarity in place of the identifiers'. This gives rise to a recursive process, where feature extraction is interleaved with multiple rounds of clustering. For this recursive process to be well-founded, we impose a topological ordering on expressions based on their dependencies (the expressions they reference). This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and several mutually-recursive expressions may appear at the same level.

\subsubsection{Algorithm}
\label{sec:algorithm}

\iffalse TODO: Not sure if you want to give a more conceptual explanation of the algorithm than the code in this figure? \fi

\iffalse TODO: Connect this to figure 1 in your explanations \fi

\iffalse TODO:
 1: Tree representation
 2. Feature vectors vs feature values
 3. Recurrence
 4. k-means algorithm
\fi

@@

    recurrentCluster :: [[(Global, Expr)]] -> [[Global]]
    recurrentCluster = go ([],[])
      where go (fs, db) []       = db
            go (fs, db) (es:ess) = let fs' = fs ++ map (extract db) es
                                       db' = kMeans fs'
                                    in go (fs', db') ess
            extract db (i, e) = (i, rt db e)
  \end{lstlisting}
  \caption{Feature extraction for Haskell Core. We send named expressions to \hs{recurrentCluster}, topologically sorted: if an element contains multiple \hs{(Global, Expr)} pairs, they are mutually-recursive. Syntax trees are first converted into \hs{RoseTree}s of \hs{Feature}s, these trees are converted to matrices, then flattened into feature vectors, then \hs{kMeans} clusters the feature vectors. The result is a list of clusters, naming their elements. More details are given in \S \ref{sec:algorithm}.}
  \label{fig:featureextractionalgorithm}
  \iffalse TODO: Split this up, get rid of float warning \fi
\end{figure}

Our feature extraction algorithm is given in figure \ref{fig:featureextractionalgorithm}. Here we highlight some key features:

### 2015-11-23 TransferReport/implementation.tex
More TODOs

  \item{Counterexample Generation}: As mentioned in \S \ref{sec:tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing. Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g. as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing. As demonstrated in \S \ref{sec:haskell}, a type's constructors might not be exported, in which case these methods fail. Also, the exponential complexity of such naive recursion makes these generators unusable in many cases. It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
\end{itemize}

>>>>

  \item{Counterexample Generation}: As mentioned in \S \ref{sec:quickcheck}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing. Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g. as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing. As demonstrated in \S \ref{sec:haskell}, a type's constructors might not be exported, in which case these methods fail. Also, the exponential complexity of such naive recursion makes these generators unusable in many cases. It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
\end{itemize}
